Linear stability of magnetized massive protoplanetary disks

### 2014-04-05 paper/summary.tex
tex updates

\section{Summary and discussion}\label{summary}
In this paper, we have performed axisymmetric linear stability
calculations of magnetized, self-gravitating disks in the local
approximation. Our models include resistivity and azimuthal fields.  
We have identified regimes under which the magneto-rotational
instability (MRI) is affected by disk self-gravity (SG).  
%what have we extended?

For a vertical field, the requirement for the MRI to operate is that its
vertical wavelength $\lambda \lesssim H$. While $\lambda$ is independent of
the strength of self-gravity $Q$, the disk thickness $H=H(Q)$ %lambda
%depends on omega, which might be
                                %lowered in a massive disk (sub-keplerian)
decreases with increasing SG, which reduces the MRI growth rates.   
%This
%effect becomes noticeable when the disk is marginally stable to GI in
%3D ($Q\sim 0.2$ or $Q_\mathrm{2D}\sim 0.7$).  
Thus, a sufficiently massive disk can surpress the MRI. %but GI may
                                %develop  instead
The MRI is also restricted to larger radial
scales as $Q$ is lowered. This means that the MRI becomes more global
in self-gravitating disks. Finally, the condition $\lambda < H$ 
requries $f(Q)/\sqrt{\beta}$ to be sufficiently small, so weaker
fields are required to permit the MRI with increasing self-gravity.

%need larger boxes 
%This suggests
%that small-scale MRI turbulence may not develop in self-gravitating
%disks.

%correlation

Interestingly, for layered resistivity we do not find layered
perturbations when the disk is massive. This is 
consistent with MRI becoming vertically global with increasing self-gravity.    
For non-self-gravitating disks $\lambda\ll H$, so the MRI can be
restricted to regions of size $L<H$, i.e. an active layer. This is not
possible if $\lambda \sim H$, as found for massive disks. Hence we
find magnetic perturbations penetrate into the high-resistivity dead
zone (e.g. $Q=0.2$ in Fig. \ref{poly_layer}), and there is no distinct
boundary between active and dead layers. This suggests that the
picture of layered accretion  \citep[e.g.][]{fleming03} may not be applicable 
to self-gravitating disks.  %strong magnetization

%SG through 
We find MRI perturbations with radial scale of $\sim H$ can acquire  
density perturbations in a massive but Toomre-stable disk. 
This occurs when the MRI is weak, for example with a strong field or
high resisitivty. We argue in that case $\lambda\sim H$, so the density
perturbation caused by the MRI can be enhanced by self-gravity.

%further enhancing density perturbations. 
At this point it is worth mentioning previous non-linear simulations
of magnetized self-gravitating galactic and circumstellar disks
\citep{kim03,fromang04a,fromang04}. These authors find self-gravity
does not enhance MRI density flucations significantly. However, they
employed ideal MHD simulations with gas-to-magnetic pressure ratios of
order $10^2$ to $10^3$. According  
to our results, self-gravity is not expected to affect the MRI in this
regime of $\beta$. On the other hand, \cite{fromang04} found
self-gravity makes MRI turbulence more coherent. This may be related
to our finding that small radial scale MRI is surpressed when
self-gravity is included in the background equilibria.

@@

%We suggest that in
%order for self-gravity to be significant,
%We suggest that in
%order for self-gravity to influence MRI,      
%cite{kim03} and \cite{fromang04}
%Numerical experiments
%varying the field strength and resisitivity indicate this is again a
%result

@@

%This is
%consistent with the MRI only affected by self-gravity for strong
%fields 
%mode classificication ambgious 
%This is consistent /expected
%numerical simulations should measure a continous range of modes
%This is
%consistent self-gravity only affecting the MRI when it is weak.  
%mag field only affect GI through pressure
%mri influenced by GI at large kx - > small scale -> need resolution

@@

%lesure 14 (strong azimuthal field)
%implications non-linear MRI simulations with a strong toroidal field
%would over-estimate growth rates (and perhaps turbulence strengths)
%if SG is not accounted for. not applicable to self-gravitating disks     
%previous linear calcs show compressibility stabilizes MRI 
%additional pressure stabilizing MRI

@@

%
%This
%prevents consideration of super-thermal toroidal fields
%\citep{pessah05}.

%tension stabilizes GI  
%This prevents us from considering super-thermal
%toroidal fields.  
%go to cylindrical disk geometry (fu & lai) 
%tension stabilizes GI

%not looked at how MRI affects GI -> it doesn't because GI isn't
%concerned with shear. only magnetic pressure. maybe a plot of GI with
%magnetic perturbations?

%spiral arms, GI angular mometum distribution 
%global modes 
%compressible term in y-induction equation 
\emph{Beyond axisymmetry.} Axisymmetric perturbations, as we have
assumed, preclude gravitational torques \citep{lynden-bell72}. 
%In 
%order to investigate how angular momentum transport 
The local non-axisymmetric stability of magnetized self-gravitating
thin disks has been considered by several authors
\citep{elmegreen87,gammie96,fan97,kim01}. However, two-dimensional models
exclude the MRI. It will be neccessary to generalize these studies to
3D in order to investigate the impact of the MRI on angular momentum
transport by gravitational instability. Furthermore, self-gravitating
disks can develop global spiral instabilities while stable against local 
axisymmetric perturbations \citep{papaloizou89,papaloizou91}. Global
non-axisymmetric linear models will be desirable to support non-linar
simulations of this kind \citep{fromang04c,fromang05}.

%Our linear problem assumes a non-zero
%vertical field is always present. It is of interest to 
%consider the opposite limit of purely toroidal fields. Of course, one
%must consider the non-axisymmetric MRI \cite{}
%The local non-axisymmetric stability of magnetized, self-gravitating
%disks has been treated in the shearing sheet framework \citep{} 
%framework by decomposing
%perturbations into shearing waves %ref.  
%
%global angular momentum exchange
%It is necessary to generalize the current models to non-axisymmetric
%perturbations. 
%It would be interesting to examine global spiral instabilities in a 
%magnetized disk. 
%In particular, to examine global spiral instabilities
%in a magnetized disk. 
%fu lai

>>>>

\section{Summary and discussion}\label{summary}
In this paper, we have performed axisymmetric linear stability
calculations of magnetized, self-gravitating, vertically stratified
disks in the local approximation. Our models include resistivity and
azimuthal fields. We have identified regimes under which the
magneto-rotational instability (MRI) is affected by disk self-gravity
(SG).

For a vertical field, the requirement for the MRI is that   
its vertical wavelength $\lambda \lesssim H$. While $\lambda$ is independent of
the strength of self-gravity $Q$, the disk thickness $H=H(Q)$ 
decreases with increasing SG, which can reducie MRI growth rates. 
Thus, a sufficiently massive disk can potentially surpress the MRI. 
The MRI is also restricted to larger radial
scales as $Q$ is lowered. This means that the MRI becomes more global
in self-gravitating disks. Finally, the condition $\lambda < H$ 
requries $f(Q)/\sqrt{\beta}$ to be sufficiently small, so weaker
fields are required to permit the MRI with increasing self-gravity.

Interestingly, for layered resistivity we do not find layered
perturbations when the disk is massive. This is 
consistent with MRI becoming vertically global with increasing self-gravity.    
For non-self-gravitating disks $\lambda\ll H$, so the MRI can be
restricted to regions of size $L<H$, i.e. an active layer. This is not
possible if $\lambda \sim H$, as found for massive disks. Hence we
find magnetic perturbations penetrate into the high-resistivity dead
zone (e.g. $Q=0.2$ in Fig. \ref{poly_layer}), and there is no distinct
boundary between active and dead layers. This suggests that the
picture of layered accretion  \citep[e.g.][]{fleming03} may not be applicable 
to self-gravitating disks.

%SG through 
We find MRI modes with radial scale of $\sim H$ can acquire  
density perturbations in massive but Toomre-stable disks. 
This occurs when the MRI is weak, for example with a strong field or
high resisitivty. We argue in that case $\lambda\sim H$, so the density
perturbation caused by the MRI can be enhanced by self-gravity.
%Our 
%interpretation involves the MRI to provide a seed density
%perturbation %in the absence of self-gravity.

%what is required? a seed pert, plus amplification

%further enhancing density perturbations. 
At this point it is worth mentioning previous non-linear simulations 
of magnetized self-gravitating galactic and circumstellar disks
\citep{kim03,fromang04a,fromang04}. These authors find self-gravity
does not enhance MRI density flucations significantly. However, they
employed ideal MHD simulations with gas-to-magnetic pressure ratios of
order $10^2$ to $10^3$. %MRI is basically incompressible 
This is qualitatively consistent with our results, as 
self-gravity is not expected to influence the MRI in this 
regime of $\beta$. On the other hand, \cite{fromang04} found
self-gravity makes MRI turbulence more coherent. This may be related
to our finding that small radial scale MRI is surpressed when
self-gravity is included in the background equilibria.

@@

@@

@@

@@

\emph{Beyond axisymmetry.} Axisymmetric perturbations, as we have
assumed, preclude gravitational torques \citep{lynden-bell72}. 
The local non-axisymmetric stability of magnetized self-gravitating
thin disks has been considered by several authors
\citep{elmegreen87,gammie96b,fan97,kim01}. However, two-dimensional models
exclude the MRI. It will be neccessary to generalize these studies to
3D in order to investigate the impact of the MRI on angular momentum
transport by gravitational instability. Furthermore, self-gravitating
disks can develop global spiral instabilities while stable against local 
axisymmetric perturbations \citep{papaloizou89,papaloizou91}. Global
non-axisymmetric linear models will be desirable to support non-linar
simulations of this kind \citep{fromang04c,fromang05}.
